### 2021-10-19 revision/ch5langevin.tex
completely done for sure.

\section{Summary}
\label{sec:langconclusion}
This chapter has presented a Langevin description of individual bedload particles saltating downstream through episodic collisions.
The model suggests that episodic particle-bed collisions control the shape of the particle velocity distribution, and it is the first model to describe the range of bedload velocity distributions which have to date been reported in experimental studies.

@@

>>>>

\section{Conclusion}
\label{sec:langconclusion}
This chapter has presented a Langevin description of individual bedload particles saltating downstream through episodic collisions.
The model suggests that episodic particle-bed collisions control the shape of the particle velocity distribution, and it is the first model to describe the range of bedload velocity distributions which have to date been reported in experimental studies.

@@

\section{Derivation of master equation}
\label{sec:langmasterderiv}
To derive the master equation from Eq. \ref{eq:langevin}, I temporarily consider the Gaussian white noise (GWN) $\xi(t)$ as a Poisson jump process having rate $r$ and jumps $ \sqrt{2 \tilde{D}} h$ with $h$ distributed as $f(h)$. I will later take a GWN limit from this noise. Similar derivations are available in some references \citep[e.g.][]{Kanazawa2017,Suweis2011}, and the precise connection between GWN and Poisson noise is discussed in \citet{VanDenBroeck1983}.

From this starting point, integrating Eq. \ref{eq:langevin} over a small time interval $\delta t$ under the It\^{o} interpretation for the collision term provides
\be     
u(t+\delta t) =
\begin{cases}
	u(t) + \tilde{\Gamma} \delta t & \text{with probability } 1- r \delta t - \nu \delta t\\
	u(t) + \sqrt{2 \tilde{D}} h &  \text{with probability } r \delta t \\
	\ve u(t) &  \text{with probability } \nu \delta t
\end{cases}.
\ee

Considering the probability $P(u,t+\delta t)$ as a sum over possible paths from $P(u,t)$ develops 
\begin{align} 
\begin{split}
P(u,t+\delta t) =
	\hspace{4pt}&(1- r \delta t - \nu \delta t)\int_{-\infty}^\infty dw  \delta(u-w-\tilde{\Gamma} \delta t) P(w,t)\\ 
	&+  r \delta t \int_{-\infty}^\infty dw \int_{-\infty}^\infty dh f(h) \delta (u - w - \sqrt{2 \tilde{D}}h) P(w,t) \\ 
	&+ \nu \delta t \int_{-\infty}^\infty dw \int_0^1 d\ve \rho(\ve)  \delta(u-w\ve) P(w,t).
\end{split}
\end{align}
Evaluating all integrals over $\delta$-functions provides 
\begin{align}
\begin{split}
 P(u,t+\delta t) = \hspace{4pt}&(1-r \delta t - \nu \delta t)P(u-\tilde{\Gamma} \delta t, t) \\&+ r \delta t\int_{-\infty}^\infty dh f(h) P(u+\sqrt{2 \tilde{D}}h) \\&+ \nu \delta t \int_0^1 \frac{ \tilde{D}\ve}{\ve} \rho(\ve) P\big(\frac{u}{\ve}\big).
\end{split}
\end{align}
Finally, taking $\delta t \rightarrow 0$ and limiting the Poisson noise involving $\sqrt{2 \tilde{D}}$ to a Gaussian white noise by taking $r \rightarrow \infty$ as $h \rightarrow 0$ such that $h^2 r = 1$ obtains the master equation \ref{eq:master}.
\section{Derivation of steady-state solution}
\label{sec:langsteadyderiv}
Defining $\tilde{P}(s) = \int_{-\infty}^\infty du e^{i u s} P(u) $ as the Fourier transform (FT) of $P(u)$ and taking the FT of Eq. \ref{eq:master} develops the recursion relation
\be \tilde{P}(s) = \frac{\tilde{P}(s\ve)}{q(s)}. \ee
where
\be q(z) = \tilde{D} z^2 - i \tilde{\Gamma} z + 1 . \label{eq:q} \ee
Recursing $N+1$ times provides
\be \tilde{P}(s) = \frac{\tilde{P}(s \ve^{N+1})}{q(s \ve^0)q(s\ve^1)\dots q(s\ve^N)}.\label{eq:recursion}\ee
The polynomials $q(z)$ can always be factored as $q(z) = \tilde{D}(z - i\lambda_-)(z - i\lambda_+)$ where
\be \lambda_\pm = \frac{\tilde{\Gamma}}{2\tilde{D}}\Big[1 \pm \sqrt{ 1 + 4\tilde{D}/\tilde{\Gamma}^2} \Big]. \label{eq:lambdas}\ee
Using these factors to expand $\tilde{P}(s)$ in partial fractions provides
\be \tilde{P}(s)  = \tilde{P}(s \ve^{N+1}) \sum_{l=0}^N  \Big[ \frac{R_l^-}{s\ve^l - i\lambda_-} + \frac{R_l^+}{s\ve^l - i\lambda_+} \Big] ,\ee
where the coefficients $R_l^\pm$ are the residues (up to constant factors) of the product $[q(s\ve^0)\dots q(s\ve^N)]^{-1}$:
\be R_l^\pm = \frac{s \ve^{l} - i\lambda_\pm }{q(s\ve^0)\dots q(s \ve^N)} \Big|_{s= i \lambda_\pm \ve^{-l}}.\ee
The Fourier-transformed Eq. \ref{eq:recursion} has a convenient feature as $N\rightarrow \infty$: since $0<\ve <1$, the prefactor $\tilde{P}( s \ve^{N+1})$ becomes the normalization condition $\tilde{P}(0)=1$ for the probability distribution $P(u)$.
Taking this limit and evaluating the residues provides 
\begin{multline} \tilde{P}(s) = \frac{1}{\tilde{D}(\lambda_+ - \lambda_-) \prod_{m=1}^\infty q(i \lambda_- \ve^m)} \sum_{l=0}^\infty \frac{i}{(s\ve^l-i\lambda_-)\prod_{m=1}^l q(i\lambda_- \ve^{-m})} 
	\\+ \frac{1}{\tilde{D}(\lambda_+ - \lambda_-) \prod_{m=1}^\infty q(i \lambda_+ \ve^m)} \sum_{l=0}^\infty \frac{-i}{(s\ve^l-i \lambda_+)\prod_{m=1}^l q(i\lambda_+ \ve^{-m})}. \end{multline}
Finally, the Fourier transforms are inverted term by term with contour integration (another use of the integral given in Eq. \ref{eq:contour}), and the definition of $q(z)$ from Eq. \ref{eq:q} is incorporated to obtain the steady-state solution quoted in Eq. \ref{eq:steadystate}.

\section{Calculation of the moments}
\label{sec:langmoments}
Taking Eq. \ref{eq:master}, multiplying by $u^k$, integrating over all space, and taking account of normalization of $P(u)$ provides a recursion relation for the moments: 
\be 0 = \tilde{D} k (k-1)\langle u^{k-2} \rangle + \tilde{\Gamma} k \langle u^{k-1} \rangle + (\ve^k-1)\langle u^k \rangle. \ee
$k=1$ provides the mean (given in Eq. \ref{eq:meanu})
\be \langle u \rangle =  \frac{\tilde{\Gamma}}{1-\ve} = \frac{\Gamma}{\nu(1-\ve)}, \ee
while $k=2$ provides the second moment
\be \langle u^2 \rangle = 2 \frac{\tilde{D} + \tilde{\Gamma} \langle u \rangle}{1-\ve^2}, \ee
leading to the velocity variance
\be \sigma_u^2 = \frac{2\tilde{D} + \tilde{\Gamma}^2}{1-\ve^2}\ee
given in Eq. \ref{eq:varu}.
\section{The weak collision limit}
\label{sec:langextremes}
Now I will demonstrate that weak collisions imply a Gaussian-like distribution for sediment velocities. The limit is challenging since the steady-state distribution Eq. \ref{eq:steadystate} and the moments above all diverge as $\ve \rightarrow 1$.
Following \cite{Hall1989}, this divergence suggests normalizing the distribution $P(u)$ using the scaled variable
\be z = \frac{u - \bra u \ket}{\sigma_u}, \ee
giving a scaled distribution
\be Q(z) = \sigma_u P(u).\ee
The resulting differential equation for $Q(z)$ is well-behaved in the elastic collision limit $\ve \rightarrow 1$.
Incorporating this transformation into Eq. \ref{eq:master} provides the ``normalized" master equation
\begin{multline}(1-\ve^2) \frac{\tilde{D}}{2\tilde{D} + \tilde{\Gamma}} Q''(z) - \frac{\tilde{\Gamma}\sqrt{1-\ve^2}}{\sqrt{2\tilde{D}+\tilde{\Gamma}^2}}Q'(z) - Q(z) \\+ \frac{1}{\ve} Q\Big(z + \Big[\frac{1-\ve}{\ve}z + \frac{\tilde{\Gamma}\sqrt{1-\ve^2}}{\ve\sqrt{2\tilde{D}+\tilde{\Gamma}^2}} \Big]\Big) = 0 .\end{multline}
This equation remains exact and is only a change of variables from Eq. \ref{eq:master}.
Now I approximate the equation for $\ve \rightarrow 1$ by expanding the final term to second order around $z=0$ before setting $\ve = 1$, obtaining
\be Q''(z) + z Q'(z) + Q(z) = 0, \ee
which is the classic Ornstein-Uhlenbeck Fokker-Planck equation whose solution is the standard normal distribution for $Q(z)$ \citep[e.g.][]{Gardiner1983}.
This solution provides Eq. \ref{eq:gaussian} when transformed back to the original variables $P(u)$ and $u$.
